---
title: "Introduction to HOIF: Higher-Order Influence Function Estimators for the ATE"
author: "Xingyu Chen"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to HOIF: Higher-Order Influence Function Estimators for the ATE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# Code chunks are not evaluated when this vignette is built: the default
# computation backend requires a Python runtime with u-stats / numpy /
# torch, which is not available (and should not be downloaded) on the
# machines that build this vignette.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
```

# Introduction

The **HOIF** package implements Higher-Order Influence Function (HOIF) estimators for Average Treatment Effect (ATE) estimation in causal inference. This methodology extends the doubly robust (DR) estimator—also known as the Augmented Inverse Propensity Weighted (AIPW) or Double Machine Learning (DML) estimator—by systematically accounting for higher-order bias terms.

## Background

The HOIF methodology was developed through a series of seminal works by James M. Robins and collaborators:

- Robins et al. (2008): *Higher order influence functions and minimax estimation of nonlinear functionals*
- Robins et al. (2017): *Minimax estimation of a functional on a structured high-dimensional model*
- Liu et al. (2017): *Semiparametric efficient empirical higher order influence function estimators*
- Liu & Li (2023): *New √n-consistent, numerically stable higher-order influence function estimators*

The key idea is to represent the bias of the standard DR estimator as a sum of higher-order influence functions, which can be estimated and used to debias the original estimator.

The computation of the higher-order U-statistics that arise in HOIF estimators is delegated to the [`ustats`](https://github.com/cxy0714/U-Statistics-R) package; the algorithm and its computational complexity are analyzed in:

- Chen, Zhang & Liu (2025): *On computing and the complexity of computing higher-order U-statistics, exactly*

## Key Features

- **Flexible basis expansion**: Supports B-splines, Fourier basis, or raw covariates
- **Multiple inversion methods**: Direct, nonlinear shrinkage, or corpcor regularization
- **Sample splitting support**: Optional K-fold cross-fitting between the
  estimation of the inverse weighted Gram matrix and the computation of the
  U-statistics (eHOIF; see the Sample Splitting section)
- **Dual computational backends**: Fast Python backend via the `ustats` package, or a pure R fallback
- **Any order with the Python backend** (pure R mode supports orders up to 6)

# Installation

```{r}
# From CRAN (once accepted):
# install.packages("HOIF")

# Development version from GitHub (also installs the ustats R package):
# install.packages("devtools")
devtools::install_github("cxy0714/HOIF")
```

## Setting up the Python backend

The default backend computes the higher-order U-statistics in Python via the
`ustats` package. The Python dependencies (`u-stats`, `numpy`, `torch`) are
**provisioned automatically**: with `reticulate` (>= 1.41) installed, the
first call that needs Python (e.g. the first `hoif_ate()` call) downloads a
private Python together with the required packages into a cached environment
and reuses it in later sessions. No manual setup is needed in most cases:

```{r}
library(HOIF)
# First call provisions the Python environment automatically:
# results <- hoif_ate(...)
```

If you prefer a persistent, dedicated environment, or want to control how
PyTorch is installed (CPU-only build vs. CUDA), use the helpers from
`ustats`:

```{r}
ustats::setup_ustats()             # CPU-only PyTorch (small download)
ustats::setup_ustats(gpu = TRUE)   # default PyPI PyTorch (CUDA on Linux)
ustats::check_ustats_setup()       # verify the environment
```

You can also point `reticulate` to your own Python/conda environment (with
`u-stats` installed via `pip install u-stats`) **before** Python initializes:

```{r}
reticulate::use_condaenv("your_env_name", required = TRUE)
# or set the RETICULATE_PYTHON environment variable
```

See `vignette("ustats", package = "ustats")` for a complete installation
guide and troubleshooting tips.

## No Python at all?

Set `pure_R_code = TRUE` in `hoif_ate()`: the U-statistics are then computed
with a pure R implementation (orders up to 6) and no Python runtime is
required.

# Quick Start Example

Let's demonstrate the basic usage with a simple simulated example.

## Generate Simulated Data

```{r}
library(HOIF)

set.seed(123)
n <- 500
p <- 2

# Generate covariates
X <- matrix(rnorm(n * p), ncol = p)

# True propensity score
true_pi <- plogis(0.5 * X[,1] - 0.3 * X[,2])

# Generate treatment
A <- rbinom(n, 1, true_pi)

# True outcome functions
mu1_true <- 1 + 0.5 * X[,1] + 0.3 * X[,2]^2
mu0_true <- 0.3 * X[,1] - 0.2 * X[,2]

# Generate outcomes
Y <- A * mu1_true + (1 - A) * mu0_true + rnorm(n, 0, 0.5)

# True ATE
true_ate <- mean(mu1_true - mu0_true)
cat("True ATE:", round(true_ate, 4), "\n")
#> True ATE: 1.3124
```

## Estimate Nuisance Functions

The package **only consumes the predicted values** of the nuisance functions;
in practice they would be estimated with flexible machine learning methods on
a **separate, independent sample** (see the Sample Splitting section below).

```{r}
# For this example, we'll use simple linear models
# (In practice, use cross-fitting with flexible ML methods)

fit_mu1 <- lm(Y ~ X, subset = A == 1)
fit_mu0 <- lm(Y ~ X, subset = A == 0)
fit_pi <- glm(A ~ X, family = binomial)

# Predict nuisance functions
mu1_hat <- predict(fit_mu1, newdata = data.frame(X = X))
mu0_hat <- predict(fit_mu0, newdata = data.frame(X = X))
pi_hat <- predict(fit_pi, type = "response")

# Ensure propensity scores are bounded away from 0 and 1
pi_hat <- pmax(pmin(pi_hat, 0.95), 0.05)
```

## Compute HOIF Estimators

```{r}
# Compute HOIF estimators without sample splitting
results <- hoif_ate(
  X = X,
  A = A,
  Y = Y,
  mu1 = mu1_hat,
  mu0 = mu0_hat,
  pi = pi_hat,
  transform_method = "none",  # Use raw covariates
  inverse_method = "direct",
  m = 5,                       # Compute up to 5th order
  sample_split = FALSE,
  backend = "torch"            # Use Python backend
)

# View results
print(results)
#> HOIF Estimators for Average Treatment Effect
#> =============================================
#>
#> Estimates by order:
#>   Order    ATE  HOIF1 HOIF0
#> 1     2 0.0015 0.0016 2e-04
#> 2     3 0.0015 0.0017 2e-04
#> 3     4 0.0015 0.0017 2e-04
#> 4     5 0.0015 0.0016 2e-04
#>
#> Final ATE estimate (highest order): 0.0015
```

Note that the reported `ATE` values are the higher-order influence function
terms of orders 2 to $m$: they estimate the **estimable bias** of the
standard first-order doubly robust (AIPW/DML) estimator and are used to
debias it. They are *not* by themselves the full ATE point estimate (compare
with the true ATE of 1.3124 above).

## Visualize Convergence

```{r}
plot(results)
```

# Main Function: `hoif_ate()`

The primary interface for computing HOIF estimators.

## Arguments

- **X**: Covariate matrix ($n \times p$)
- **A**: Treatment vector (0 or 1)
- **Y**: Outcome vector
- **mu1, mu0**: Predicted outcomes under treatment and control
- **pi**: Predicted propensity scores
- **transform_method**: Covariate transformation method
  - `"none"`: No transformation (use raw covariates)
  - `"splines"`: B-splines basis expansion
  - `"fourier"`: Fourier basis expansion
- **basis_dim**: Number of basis functions (required if using splines or Fourier)
- **inverse_method**: Method for Gram matrix inversion
  - `"direct"`: Cholesky-based inversion (Moore-Penrose fallback)
  - `"nlshrink"`: Nonlinear shrinkage (Ledoit-Wolf)
  - `"corpcor"`: Shrinkage via corpcor package
- **m**: Maximum order (any order for the Python backend, 2-6 for pure R)
- **sample_split**: Logical; whether to use K-fold cross-fitting
- **n_folds**: Number of folds for sample splitting
- **backend**: `"torch"` (default) or `"numpy"` for the Python backend
- **dtype**: Numeric precision of the Python backend (`"float32"`,
  `"float64"`, or `NULL` for automatic selection)
- **pure_R_code**: Use pure R implementation (limited to order 6)
- **seed**: Random seed for reproducibility

## Return Value

A list of class `"hoif_ate"` containing:

- **ATE**: Vector of ATE estimates for orders 2 to m
- **HOIF1**: Vector of treatment arm HOIF estimates
- **HOIF0**: Vector of control arm HOIF estimates
- **IIFF1**: Vector of treatment arm incremental influence function terms
- **IIFF0**: Vector of control arm incremental influence function terms
- **orders**: Vector of orders (2:m)
- **convergence_data**: Data frame for plotting

# Advanced Usage

## Using Basis Expansion

For more flexible modeling of covariate relationships:

```{r}
# B-splines basis
results_splines <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  transform_method = "splines",
  basis_dim = 5,        # 5 basis functions per covariate
  degree = 3,           # Cubic splines
  m = 5,
  sample_split = FALSE
)

# Fourier basis
results_fourier <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  transform_method = "fourier",
  basis_dim = 4,        # 4 Fourier components per covariate
  period = 1,
  m = 5,
  sample_split = FALSE
)
```

## Sample Splitting (Cross-Fitting)

To cross-fit the estimation of the inverse weighted Gram matrix against the
computation of the U-statistics — the eHOIF estimator (see the Sample
Splitting section under Mathematical Background for what exactly is split):

```{r}
results_cf <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  transform_method = "none",
  m = 5,
  sample_split = TRUE,
  n_folds = 5,          # 5-fold cross-fitting
  seed = 42
)
```

## Regularized Gram Matrix Inversion

For high-dimensional settings or when facing numerical instability:

```{r}
# Nonlinear shrinkage
results_nlshrink <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  inverse_method = "nlshrink",
  m = 5
)

# corpcor shrinkage
results_corpcor <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  inverse_method = "corpcor",
  m = 5
)
```

## Pure R Backend

If the Python environment is not available or for smaller problems:

```{r}
results_pure_r <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  pure_R_code = TRUE,
  m = 6  # Maximum 6 for pure R
)
```

# Mathematical Background

## The HOIF Framework

The HOIF estimator for ATE is defined as:

$$
\mathbb{ATE}_m(\hat{\Omega}^a) = \mathbb{HOIF}^1_m - \mathbb{HOIF}^0_m
$$

where for each treatment arm $a \in \{0,1\}$:

$$
\mathbb{HOIF}^a_m(\hat{\Omega}^a) = \sum_{j=2}^m \mathbb{IF}^a_j(\hat{\Omega}^a) = \sum_{i=2}^m \sum_{j=2}^{i} \binom{i-2}{i-j} \mathbb{U}^a_j(\hat{\Omega}^a)
$$

## U-Statistics Formulation

The core computational building block is the U-statistic:

$$
\mathbb{U}^a_m(\hat{\Omega}^a) = (-1)^m \frac{(n-m)!}{n!} \sum_{\substack{(i_1, \ldots, i_m) : \\ i_1 \neq \cdots \neq i_m}} r^a_{i_1} \prod_{s=1}^{m-1} \{Z_{i_s}^\top \hat{\Omega}^a s^a_{i_{s+1}} Z_{i_{s+1}}\} R^a_{i_m}
$$

where:

- $s^a_i = A_i^a (1-A_i)^{1-a}$ is the treatment indicator
- $r^a_i = 1 - s^a_i / \pi^a(X_i)$ is the inverse propensity weight residual
- $R^a_i = Y_i - \hat{\mu}(a, X_i)$ is the outcome residual
- $Z_i$ are the transformed covariates
- $\hat{\Omega}^a$ is the inverse weighted Gram matrix

The full algorithmic workflow, mathematical formulas, and all parameters are
documented in the PDF shipped with the package:
`system.file("extdoc", "HOIF.pdf", package = "HOIF")`.

## Sample Splitting

Conceptually, HOIF estimation involves **three** separate estimation tasks,
and ideally each would use its own, independent part of the data:

1. **The nuisance functions**: estimating $\hat{\mu}(1, X)$,
   $\hat{\mu}(0, X)$ and $\hat{\pi}(X)$;
2. **The inverse weighted Gram matrix** $\hat{\Omega}^a$;
3. **The higher-order U-statistics** built from $\hat{\Omega}^a$.

This package deliberately does **not** implement task 1: `hoif_ate()` only
takes the *predicted values* `mu1`, `mu0`, `pi` as inputs, and how (and on
which sample) the nuisance functions are estimated is entirely up to the
user. Accordingly, the full three-way cross-fitting is **not** performed by
the package; the `sample_split` argument only controls the split between
tasks 2 and 3:

- **`sample_split = TRUE` (eHOIF)**: the sample passed to `hoif_ate()` is
  split into `n_folds` folds $(I_1, \ldots, I_K)$. For each fold $j$,
  $\hat{\Omega}^a$ is estimated on the observations **not** in $I_j$, the
  U-statistics are computed on the observations **in** $I_j$, and the
  results are averaged across the folds.
- **`sample_split = FALSE` (sHOIF)**: $\hat{\Omega}^a$ and the U-statistics
  are computed on the same sample, without distinction.

# Computational Details

## Python Backend (ustats)

The default backend uses the [`ustats`](https://github.com/cxy0714/U-Statistics-R) R package, an interface to the Python package [`u-stats`](https://pypi.org/project/u-stats/), for efficient U-statistic computation:

- Supports PyTorch and NumPy backends (GPU acceleration with CUDA-enabled PyTorch)
- Handles arbitrary order U-statistics
- Optimized for memory and speed via Einstein summation

For HOIF estimators of the ATE, a key takeaway from Chen, Zhang & Liu (2025) is:

- when the order $m \le 7$, the computational complexity is
  $O(n^3 + nk^2 + k^3 + n^2 k)$
- when the order $m > 7$, the computational complexity exceeds
  $O(n^4 + nk^2 + k^3 + n^2 k)$

where $n$ is the sample size and $k$ is the dimension of the transformed covariates.

## Pure R Implementation

The pure R fallback (`pure_R_code = TRUE`):

- Implements orders 2-6 using recursive matrix operations
- Uses the `SMUT` package for fast matrix multiplication
- Removes diagonal contributions to ensure proper degenerate U-statistics
- Suitable for smaller datasets or when Python is unavailable

## Performance Considerations

- **Order**: Higher orders increase computation time
- **Basis dimension**: Larger basis increases memory and computation
- **Cross-fitting**: Multiplies computation by number of folds
- **Precision**: On a CUDA GPU the Python backend defaults to `float32`;
  pass `dtype = "float64"` for maximum numerical precision

# Practical Recommendations

## Choosing the Transformation Method

- **"none"**: Use the covariates as provided (e.g. when you have already
  constructed your own basis)
- **"splines"**: Good for smooth nonlinear relationships
- **"fourier"**: Suitable for periodic patterns or high-frequency features

## Choosing the Order

For orders $m \le 7$ the whole computation reduces to ordinary matrix
algebra with complexity $O(n^3 + nk^2 + k^3 + n^2 k)$ (Chen, Zhang & Liu,
2025), so going up to order 7 is cheap. A practical default is therefore to
compute all orders up to $m = 7$ (the package default) and inspect the
convergence plot. Beyond order 7 the complexity grows to at least
$O(n^4 + nk^2 + k^3 + n^2 k)$, so expect substantially longer run times.

## Use a GPU if Available

With `backend = "torch"` and a CUDA-enabled PyTorch (e.g. installed via
`ustats::setup_ustats(gpu = TRUE)`), the U-statistics are computed on the
GPU automatically — by far the most effective speedup for large $n$ or $k$.
On a GPU the default precision is `float32`; pass `dtype = "float64"` if
you need maximum numerical precision.

## Sample Splitting

See the Sample Splitting section under Mathematical Background: the package
only cross-fits the inverse weighted Gram matrix against the U-statistics
(`sample_split = TRUE`, eHOIF). The nuisance functions are inputs —
estimating them, ideally on a separate, independent sample, is the user's
responsibility.

# Troubleshooting

## Python Backend Issues

Verify the Python environment used by `ustats`:

```{r}
ustats::check_ustats_setup()
```

If the environment cannot be repaired easily, switch to the pure R implementation:

```{r}
results <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  pure_R_code = TRUE,
  m = 6  # Limited to order 6
)
```

## Numerical Instability

If you see warnings about matrix inversion:

```{r}
# Try shrinkage methods
results <- hoif_ate(
  X = X, A = A, Y = Y,
  mu1 = mu1_hat, mu0 = mu0_hat, pi = pi_hat,
  inverse_method = "nlshrink"  # or "corpcor"
)
```

## Extreme Propensity Scores

Always trim propensity scores to avoid numerical issues:

```{r}
# Trim to [0.05, 0.95]
pi_hat <- pmax(pmin(pi_hat, 0.95), 0.05)
```

# References

1. Robins, J. M., Li, L., Tchetgen Tchetgen, E., & van der Vaart, A. (2008). Higher order influence functions and minimax estimation of nonlinear functionals. *Probability and Statistics: Essays in Honor of David A. Freedman*, 335-421. <doi:10.1214/193940307000000527>

2. Robins, J. M., Li, L., Mukherjee, R., Tchetgen Tchetgen, E., & van der Vaart, A. (2017). Minimax estimation of a functional on a structured high-dimensional model. *Annals of Statistics*, 45(5), 1951-1987.

3. Liu, L., Mukherjee, R., Newey, W. K., & Robins, J. M. (2017). Semiparametric efficient empirical higher order influence function estimators. *arXiv preprint* <doi:10.48550/arXiv.1705.07577>

4. Liu, L., & Li, C. (2023). New √n-consistent, numerically stable higher-order influence function estimators. *arXiv preprint* <doi:10.48550/arXiv.2302.08097>

5. Chen, X., Zhang, R., & Liu, L. (2025). On computing and the complexity of computing higher-order U-statistics, exactly. *arXiv preprint* <doi:10.48550/arXiv.2508.12627>
